Efficient forwards-time simulation with ancestral recombination graphs

### 2017-10-20 forwards_paper.tex
Added n<<N assumption paragraph & refs.

\usepackage{color}
\newcommand{\plr}[1]{{\em \color{blue} #1}}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g. $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of (a) random mating, and (b) neutrality.
\plr{Brief statement why this is.  Also include stationarity?}
Both assumptions can be side-stepped to a limited extent, and so coalescent methods are now commonly used to
simulate the results of population dynamics of a collection of randomly mating populations exchanging migrants,
having a small number of loci under selection.
Mapping results of such models onto real species can be challenging,
as these are often distributed across geographical space and may have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
These considerations, and increasing computational power, have led to a resurgence of interest in forwards-time, individual-based simulations.

With modern computing power, pure demographic calculations are not a barrier,
even though biological population sizes are often above $10^6$,
and coalescent theory tells us that a population of size $N$ 
must be run for a multiple of $N$ generations to produce stable genetic patterns.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simluation becomes quickly prohibitive.

However, it is thought that much of that variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
one can lay down neutral mutations on top of that history afterwards,
without loss of generality.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime
to efficiently record, and later process, this history.

>>>>

\usepackage{color}
\newcommand{\plr}[1]{{\em \color{blue} #1}}
\newcommand{\jk}[1]{{\em \color{red} #1}}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g.\ $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of (a) random mating, and (b) neutrality.
\plr{Brief statement why this is.  Also include stationarity?}
Both assumptions can be side-stepped to a limited extent, and so coalescent methods are now commonly used to
simulate the results of population dynamics of a collection of randomly mating populations exchanging migrants,
having a small number of loci under selection.
Mapping results of such models onto real species can be challenging,
as these are often distributed across geographical space and may have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
These considerations, and increasing computational power, have led to a resurgence of interest in forwards-time, individual-based simulations.

Another critical assumption of the coalescent is that the sample size $n$ is much smaller
than the effective population size $N_e$. With the very large sample sizes becoming common
in human genetics, this is no longer a safe assumption. For example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
reference panels in genome wide association studies. While many results of the coalescent are surpisingly
robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
are distorted and coalescent simulations may be
misleading~\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
A hybrid approach, in which the recent history of a large sample of
individuals is simulated under a detailed forwards-in-time model and the deep history
under the coalescent, is an attractive prospect.
\jk{Not sure where this should go; perhaps in the discussion? We can make the point that
by keeping track of the genealogies forwards in time we can easily combine these with the
deep history simulated by msprime}

With modern computing power, pure demographic calculations are not a barrier,
even though biological population sizes are often above $10^6$,
and coalescent theory tells us that a population of size $N$
must be run for a multiple of $N$ generations to produce stable genetic patterns.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation becomes quickly prohibitive.

However, it is thought that much of that variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
one can lay down neutral mutations on top of that history afterwards,
without loss of generality.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime\
to efficiently record, and later process, this history.
